Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2014-12-18 parts/example.tex
'saturated' GLCs + several fixes

\modMF{
Finally, the static analysis results presented in
\pref{ssec:ua} and \pref{ssec:ordered-ua} can be used to check if the model is functional,
\ie if gene $a$ can oscillate, thus leading to the formation of stripes.
Starting from context $\ctx = \PHstate{f_1, a_0, c_0, fc_{10}, fa_{00}}$,
we thus want to check the reachability of
$\w = \PHobj{a_0}{a_1} \PHconcat \PHobj{a_1}{a_0} \PHconcat \PHobj{a_0}{a_1}$.
Rather than checking this whole objective sequence with \pref{th:approxinf},
we break it using \pref{thm:ordered-ua}.
We first build the first saturated graph of local causality
based on the first objective, $\w_1 = \PHobj{a_0}{a_1}$, in the initial context $\ctx$.
This graph is depicted in \pref{fig:metazoan-sa}(left).
%and given \pref{tmh:approxinf}, we have:
We have especially:
\[\allprocs(\thisB{\w_1}{\ctx}) = \{ a_0, c_0, f_1, fc_{10} \}\]
\[\lastprocs(\PHobj{a_0}{a_1}) = \{\{ a_1, c_0, f_1, fc_{10} \}\}\]
Therefore, the second step consists of computing the saturated graph of local causality
of the objective $\w_2 = \PHobj{a_1}{a_0}$
in the context
\[\ctx' = \ctx \Cap \allprocs(\thisB{\w_1}{\ctx}) \Cap \lastprocs(\PHobj{a_0}{a_1}) = \PHstate{a_1, c_0, f_1, fc_{10}, fa_{00}}\]
This second graph is given in \pref{fig:metazoan-sa}(middle).
Finally, because
\[\allprocs(\thisB{\w_2}{\ctx'}) = \{ a_1, c_1, f_1, fa_{11} \}\]
\[\lastprocs(\PHobj{a_1}{a_0}) = \{\{ a_0, c_1 \}\}\]
The last step is then checking 
the objective $\w_3 = \PHobj{a_0}{a_1}$
in the context
\[\ctx'' = \ctx' \Cap \allprocs(\thisB{\w_2}{\ctx'}) \Cap \lastprocs(\PHobj{a_1}{a_0}) = \PHstate{a_0, c_1, f_1, fc_{10}, fa_{11}}\]
This last graph is depicted in 
As all the produced \pref{fig:metazoan-sa}(right).
In conclusion, by using \pref{thm:ordered-sa} on these saturated graphs of local causality,
we can conclude that the initial objective $\w$ is reachable,
and thus that the PH model of \pref{fig:metazoan-ph} is functional.
This result can be extended to the model of \pref{fig:metazoan-php}
because they have the same dynamics (given \pref{th:bisimPHP}).
}

>>>>

\modMF{
Finally, the static analysis results presented in
\pref{ssec:ua} and \pref{ssec:ordered-ua} can be used to check if the model is functional,
\ie if gene $a$ can oscillate, thus leading to the formation of stripes.
Starting from context $\ctx = \PHstate{f_1, a_0, c_0, fc_{10}, fa_{00}}$,
we thus want to check the reachability of
$\w = \PHobj{a_0}{a_1} \PHconcat \PHobj{a_1}{a_0} \PHconcat \PHobj{a_0}{a_1}$.
Rather than checking this whole objective sequence with \pref{th:approxinf},
we break it using \pref{thm:ordered-ua}.
We first build the first saturated graph of local causality
based on the first objective, $\w_1 = \PHobj{a_0}{a_1}$, in the initial context $\ctx$.
This graph is depicted in \pref{fig:metazoan-sa}(left).
%and given \pref{tmh:approxinf}, we have:
We have especially:
\[\allprocs(\thisB{\w_1}{\ctx}) = \{ a_0, c_0, f_1, fc_{10} \}\]
\[\lastprocs(\PHobj{a_0}{a_1}) = \{\{ a_1, c_0, f_1, fc_{10} \}\}\]
Therefore, the second step consists of computing the saturated graph of local causality
of the objective $\w_2 = \PHobj{a_1}{a_0}$
in the context
\[\ctx' = \ctx \Cap \allprocs(\thisB{\w_1}{\ctx}) \Cap \lastprocs(\PHobj{a_0}{a_1}) = \PHstate{a_1, c_0, f_1, fc_{10}, fa_{00}}\]
This second graph is given in \pref{fig:metazoan-sa}(middle).
Finally, because
\[\allprocs(\thisB{\w_2}{\ctx'}) = \{ a_1, c_1, f_1, fa_{11} \}\]
\[\lastprocs(\PHobj{a_1}{a_0}) = \{\{ a_0, c_1 \}\}\]
The last step is then checking 
the objective $\w_3 = \PHobj{a_0}{a_1}$
in the context
\[\ctx'' = \ctx' \Cap \allprocs(\thisB{\w_2}{\ctx'}) \Cap \lastprocs(\PHobj{a_1}{a_0}) = \PHstate{a_0, c_1, f_1, fc_{10}, fa_{11}}\]
This last graph is depicted in 
As all the produced \pref{fig:metazoan-sa}(right).
In conclusion, by using \pref{thm:ordered-ua} on these saturated graphs of local causality,
we can conclude that the initial objective $\w$ is reachable,
and thus that the PH model of \pref{fig:metazoan-ph} is functional.
This result can be extended to the model of \pref{fig:metazoan-php}
because they have the same dynamics (given \pref{th:bisimPHP}).
}

### 2014-12-18 parts/sa.tex
'saturated' GLCs + several fixes

\begin{definition}
\label{def:glc}
  The graph of local causality $\cwB \DEF (\Bv, \Be)$ is defined as: $\cwB \DEF \lfp{\aB^\w_\ctx}{\myB}{\aB^\w_{\ctx \Cap \allprocs(\myB)}}$,
  where $\myB \DEF (\cwV, \cwE)$ is the smallest graph with
  $\cwV \subseteq \Proc \cup \Obj \cup \Sol$ and $\cwE \subseteq \cwV \times \cwV$
  so that:
  \begin{align}
    \w &\subseteq \cwV \label{eq:Vw} \\
%    P \in \cwVObj &\Rightarrow \PHbounce(P) \in \cwV \label{eq:Vproc} \\
    (x, y) \in \cwE &\Rightarrow y \in \cwV \label{eq:VE} \\
    P \in \cwVObj \wedge ps \in \modMF{\aBS}(P) &\Rightarrow (P, ps) \in \cwE \label{eq:ESol1} \\
    ps \in \cwVSol \wedge a_i \in ps &\Rightarrow (ps, a_i) \in \cwE \label{eq:ESol2} \\
    a \in \components \wedge a_i \in \cwVProc \wedge a_j \in \ctx &\Rightarrow (a_i, \PHobjp{a}{j}{i}) \in \cwE \label{eq:EReq} \\
    a \in \cs \wedge a_i \in \cwVProc \wedge ps \in \csState(a_i) &\Rightarrow (a_i, ps) \in \cwE \label{eq:EPrio} \\
    P \in \cwVObj \wedge q \in \gCont_\ctx(\PHsort(P), P) &\Rightarrow (P, \PHobj{q}{\PHbounce(P)}) \in \cwE \label{eq:ECont} \!
  \end{align}
\end{definition}

@@

In the saturated graph of local causality, an edge $(p, ps) \in \Proc \times \Sol$
is said to be coherent (\pref{def:coherent}) if none of the processes in $ps$ conflict with the children processes of $ps$.
Then, \pref{th:approxinf} gives a sufficient condition for the concretisation of a sequence of objectives in a given context,
which is derived immediately from the graph of local causality.
A proof of this theorem is given in \pref{suppl:demoapproxinf}.
\begin{definition}[Coherent edge]
\label{def:coherent}
  An edge $(x, y) \in \cwE$ is said to be coherent if and only if:
  $(x, y) \in \Be \cap (\Proc \times \Sol) \Rightarrow y$ has no children process $a_j \in \Bv \cap \Proc$ so that $\exists a_i \in y$, $a_i \neq a_j$.
\end{definition}

@@

Computing the graph of local causality is polynomial in the number of sorts in $\PH$ and exponential in the number of processes in one sort.
Checking the properties allowing to apply \pref{th:approxinf} is polynomial in the size of the graph.
Furthermore, it is possible to compute only a subset of $\V \cap \Sol$;
in this case, the overall method turns out to be exponential in the number of solutions to each objective.
Our method can thus be considered as efficient compared to regular model-checking which is usually PSPACE-complete~\cite{Harel02}.

\begin{example}
  Let $\PH' = (\PHs, \PHl, \PHh'^{\langle 1 \rangle})$ be the “merged” version of the PH in \pref{fig:ph-livelock}, that is: $\PHh'^{\langle 1 \rangle} = \PHh^{(1)} \cup \PHh^{(2)}$,
  which is how it would have been represented in the semantics without priorities.
  Due to spurious behaviours inherent to the cooperative sorts in this semantics, 
  the original under-approximation developed in~\cite{PMR12-MSCS} concludes that $c_1$ is reachable in $\PH'$ from $\ctx = \PHstate{a_1, b_0, c_0, ab_{10}}$.

  Such unwanted behaviours are palliated by the semantics of PH with priorities proposed in this paper.
  Indeed, the under-approximation given in \pref{th:approxinf} does not conclude regarding the reachability of $c_1$,
  as one of the edges of the resulting graph of local causality is not coherent (\pref{def:coherent}),
  as shown in in \pref{fig:sa-livelock}.
  (However, from the inconclusiveness of \pref{th:approxinf}, one cannot conclude about the unreachability of $c_1$.
  Such analysis should be driven for instance with over-approximation methods developed in~\cite{PMR12-MSCS}.)

  However, if $\PHhit{a_0}{b_0}{b_1}$ and $\PHhit{b_0}{a_0}{a_1}$ are replaced by the auto-actions
  $\PHhit{a_0}{a_0}{a_1}$ and $\PHhit{b_0}{b_0}{b_1}$,
  then \pref{th:approxinf} concludes that $c_1$ is reachable from $\ctx$.

>>>>

\begin{definition}
\label{def:glc}
  The \modMF{saturated} graph of local causality $\cwB \DEF (\Bv, \Be)$ is defined as:
  $\cwB \DEF \lfp{\aB^\w_\ctx}{\myB}{\aB^\w_{\ctx \Cap \allprocs(\myB)}}$,
  where $\myB \DEF (\cwV, \cwE)$ is the smallest graph with
  $\cwV \subseteq \Proc \cup \Obj \cup \Sol$ and $\cwE \subseteq \cwV \times \cwV$
  so that:
  \begin{align}
    \w &\subseteq \cwV \label{eq:Vw} \\
%    P \in \cwVObj &\Rightarrow \PHbounce(P) \in \cwV \label{eq:Vproc} \\
    (x, y) \in \cwE &\Rightarrow y \in \cwV \label{eq:VE} \\
    P \in \cwVObj \wedge ps \in \modMF{\aBS}(P) &\Rightarrow (P, ps) \in \cwE \label{eq:ESol1} \\
    ps \in \cwVSol \wedge a_i \in ps &\Rightarrow (ps, a_i) \in \cwE \label{eq:ESol2} \\
    a \in \components \wedge a_i \in \cwVProc \wedge a_j \in \ctx &\Rightarrow (a_i, \PHobjp{a}{j}{i}) \in \cwE \label{eq:EReq} \\
    a \in \cs \wedge a_i \in \cwVProc \wedge ps \in \csState(a_i) &\Rightarrow (a_i, ps) \in \cwE \label{eq:EPrio} \\
    P \in \cwVObj \wedge q \in \gCont_\ctx(\PHsort(P), P) &\Rightarrow (P, \PHobj{q}{\PHbounce(P)}) \in \cwE \label{eq:ECont} \!
  \end{align}
\end{definition}

@@

In the saturated graph of local causality, an edge $(p, ps) \in \Proc \times \Sol$
is said to be coherent (\pref{def:coherent}) if none of the processes in $ps$ conflict with the children processes of $ps$.
Then, \pref{th:approxinf} gives a sufficient condition for the concretisation of a sequence of objectives in a given context,
which is derived immediately from the saturated graph of local causality.
A proof of this theorem is given in \pref{suppl:demoapproxinf}.
\begin{definition}[Coherent edge]
\label{def:coherent}
  An edge $(x, y) \in \cwE$ is said to be coherent if and only if:
  $(x, y) \in \Be \cap (\Proc \times \Sol) \Rightarrow y$ has no children process $a_j \in \Bv \cap \Proc$ so that $\exists a_i \in y$, $a_i \neq a_j$.
\end{definition}

@@

Computing the saturated graph of local causality is polynomial in the number of sorts in $\PH$ and exponential in the number of processes in one sort.
Checking the properties allowing to apply \pref{th:approxinf} is polynomial in the size of the graph.
Furthermore, it is possible to compute only a subset of $\V \cap \Sol$;
in this case, the overall method turns out to be exponential in the number of solutions to each objective.
Our method can thus be considered as efficient compared to regular model-checking which is usually PSPACE-complete~\cite{Harel02}.

\begin{example}
  Let $\PH' = (\PHs, \PHl, \PHh'^{\langle 1 \rangle})$ be the “merged” version of the PH
  depicted in \pref{fig:ph-livelock}, as it was already defined in \pref{ssec:priorties},
  that is: $\PHh'^{\langle 1 \rangle} = \PHh^{(1)} \cup \PHh^{(2)}$.
  In other words,
  $\PH'$ is how $\PH$ would have been represented in the semantics without priorities.
  We showed in \pref{ssec:priorties} that,
  due to spurious behaviours inherent to the cooperative sorts in this semantics,
  process $c_1$ is reachable
  from the initial context $\ctx = \PHstate{a_1, b_0, c_0, ab_{10}}$.
%   This is confirmed by the original under-approximation
%   developed in~\cite{PMR12-MSCS}, which gives the same conclusion.

  Such unwanted behaviours are palliated by the semantics of PH with priorities proposed in this paper.
  Indeed, the under-approximation given in \pref{th:approxinf} does not conclude regarding the reachability of $c_1$,
  as one of the edges of the resulting saturated graph of local causality is not coherent (\pref{def:coherent}),
  as shown in in \pref{fig:sa-livelock}.
  (However, from the inconclusiveness of \pref{th:approxinf}, one cannot conclude about the unreachability of $c_1$.
  Such analysis should be driven for instance with over-approximation methods developed in~\cite{PMR12-MSCS}.)

  We note however that,
  if $\PHhit{a_0}{b_0}{b_1}$ and $\PHhit{b_0}{a_0}{a_1}$ are replaced by the auto-actions
  $\PHhit{a_0}{a_0}{a_1}$ and $\PHhit{b_0}{b_0}{b_1}$,
  then the resulting saturated graph of local causality changes, and
  \pref{th:approxinf} concludes that $c_1$ is then reachable from $\ctx$.
